---
title: "Common names for Fish4life"
author: "Lasse Rauert"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 4
    toc_float: true
---

# Libraries
```{r, results="hide", message=FALSE}
library(rgbif)

library(dplyr)
library(tidyr)

library(purrr)
```
# Load name list data
```{r}
f4l <- read.csv("../data/names_Fish4life.csv")
f4l
```
# Match fish_names to GBIF Backbone --> full parsed taxonomy
```{r}
fish_names_matched<-name_backbone_checklist(f4l$Scientific.Name)
fish_names_matched
write.csv(fish_names_matched,"../data/sci_names.csv", row.names = FALSE)
```


# add common names from GBIF Taxonomic backbone to the f4l Table
## function to get common names for singel species
```{r}
get_common_names <- function(species_name) {
  key_info <- name_backbone(name = species_name)
  
  if (!"speciesKey" %in% names(key_info)) {
    return(data.frame(Scientific.Name = species_name))
  }
  
  key <- key_info$speciesKey
  v_names <- name_usage(key = key, data = 'vernacularNames')
  
  if (length(v_names$data) == 0 || nrow(v_names$data) == 0 || !"language" %in% names(v_names$data)) {
    return(data.frame(Scientific.Name = species_name))
  }
  
  clean_v_data <- v_names$data %>%
    filter(!is.na(language) & language != "") %>%
    group_by(language) %>%
    summarise(vernacularName = first(vernacularName), .groups = 'drop') %>%
    mutate(Scientific.Name = species_name)
  
  return(clean_v_data)
}
```

## apply function to dataset
```{r}
unique_species <- unique(f4l$Scientific.Name)
all_common_names <- map_dfr(unique_species, get_common_names)
```

## join data
```{r}
# pivot 
wide_common_names <- all_common_names %>%
  pivot_wider(names_from = language, values_from = vernacularName)

# join
final_f4l <- f4l %>%
  left_join(wide_common_names, by = "Scientific.Name")

head(final_f4l)
```

```{r}
write.csv(final_f4l, "../data/Fish4life_w_common.csv", row.names = FALSE)
```
# info / citations

```{r}
library(grateful)
cite_packages(out.dir = ".")
```


We used R v. 4.5.1 (R Core Team 2025) and the following R packages: rgbif v. 3.8.3 (Chamberlain and Boettiger 2017; Chamberlain et al. 2026), rmarkdown v. 2.29 (Xie, Allaire, and Grolemund 2018; Xie, Dervieux, and Riederer 2020; Allaire et al. 2024), tidyverse v. 2.0.0 (Wickham et al. 2019).


## Package citations
```
Allaire, JJ, Yihui Xie, Christophe Dervieux, Jonathan McPherson, Javier Luraschi, Kevin Ushey, Aron Atkins, et al. 2024. rmarkdown: Dynamic Documents for r. https://github.com/rstudio/rmarkdown.

Chamberlain, Scott, Vijay Barve, Dan Mcglinn, Damiano Oldoni, Peter Desmet, Laurens Geffert, and Karthik Ram. 2026. rgbif: Interface to the Global Biodiversity Information Facility API. https://CRAN.R-project.org/package=rgbif.

Chamberlain, Scott, and Carl Boettiger. 2017. “R Python, and Ruby Clients for GBIF Species Occurrence Data.” PeerJ PrePrints. https://doi.org/10.7287/peerj.preprints.3304v1.

R Core Team. 2025. R: A Language and Environment for Statistical Computing. Vienna, Austria: R Foundation for Statistical Computing. https://www.R-project.org/.

Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D’Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. “Welcome to the tidyverse.” Journal of Open Source Software 4 (43): 1686. https://doi.org/10.21105/joss.01686.
Xie, Yihui, J. J. Allaire, and Garrett Grolemund. 2018. R Markdown: The Definitive Guide. Boca Raton, Florida: Chapman; Hall/CRC. https://bookdown.org/yihui/rmarkdown.

Xie, Yihui, Christophe Dervieux, and Emily Riederer. 2020. R Markdown Cookbook. Boca Raton, Florida: Chapman; Hall/CRC. https://bookdown.org/yihui/rmarkdown-cookbook.
```
```{r}
sessionInfo()
```